Application of Network Flow and Linear Programming to Scheduling of Clinicians

### 2019-08-21 paper/paper.discussion.tex
Rename divisions -> services throughout
Also additional changes to results section following comments

Our formulation includes hard constraints to ensure the schedule satisfies hospital and logistics requirements. It also aims to satisfy the work preferences of clinicians in a clinical department by optimizing a multi-goal objective function. We are able to use the tool to generate schedules for a wide variety of simulated departments, supporting upwards of 50 total clinicians. Moreover, our formulation is able to accommodate the requests of clinicians without negatively affecting the runtime. The scenarios when the ILP formulation starts to have trouble are ones with multiple services offered within a single clinical department. Once there are more than 10 clinicians in total, providing services in 2 or more divisions, the constraint space becomes very complex, making it difficult for the tool to find an optimal solution in a reasonable amount of time. For these cases, we recommend allowing consecutive blocks, to simplify the problem and find a solution much faster. \\

>>>>

Our formulation includes hard constraints to ensure the schedule satisfies hospital and logistics requirements. It also aims to satisfy the work preferences of clinicians in a clinical department by optimizing a multi-goal objective function. We are able to use the tool to generate schedules for a wide variety of simulated departments, supporting upwards of 50 total clinicians. Moreover, our formulation is able to accommodate the requests of clinicians without negatively affecting the runtime. The scenarios when the ILP formulation starts to have trouble are ones with multiple services offered within a single clinical department. Once there are more than 10 clinicians in total, providing 2 or more services, the search space becomes very complex, making it difficult for the tool to find an optimal solution in a reasonable amount of time. For these cases, we recommend allowing consecutive blocks, to simplify the problem and find a solution much faster. \\

### 2019-08-21 paper/paper.problem.tex
Rename divisions -> services throughout
Also additional changes to results section following comments

\begin{table}[h]
	\centering
	\begin{tabular}{ l l l }
		\hline
		\textbf{Constraint Name} & \textbf{Description}                                                                                                           & \textbf{Type} \\ \hline
		Block Coverage           & \makecell[l]{each division needs to have exactly \\ one clinician that covers any given block}                                 & Hard          \\ \hline
		Weekend Coverage         & \makecell[l]{every weekend needs to have exactly \\ one clinician that covers it}                                              & Hard          \\ \hline
		Min/Max                  & \makecell[l]{in a given division, each clinician can only \\ work between the minimum and maximum \\ number of allowed blocks} & Hard          \\ \hline
		No Consecutive Blocks    & \makecell[l]{any clinician should not work \\ two consecutive blocks, across all divisions}                                    & Hard          \\ \hline
		No Consecutive Weekends  & \makecell[l]{any clinician should not work two consecutive \\ weekends}                                                        & Hard          \\ \hline
		Equal Weekends           & \makecell[l]{weekends should be equally distributed \\ between clinicians}                                                     & Hard          \\ \hline
		Equal Holidays           & \makecell[l]{long weekends should be equally distributed \\ between clinicians}                                                & Hard          \\ \hline
		Block Requests           & \makecell[l]{each clinician can request to be off service \\ during certain blocks throughout the year}                        & Soft          \\ \hline
		Weekend Requests         & \makecell[l]{each clinician can request to be off service \\ during certain weekends throughout the year}                      & Soft          \\ \hline
		Block-Weekend Adjacency  & \makecell[l]{the block and weekend assignments of a given \\ clinician should be adjacent}                                     & Soft
	\end{tabular}
	\caption{Summary of the constraints for the clinician scheduling problem}
	\label{tbl:constraint-summary}
\end{table}

>>>>

\begin{table}[h]
	\centering
	\begin{tabular}{ l l l }
		\hline
		\textbf{Constraint Name} & \textbf{Description}                                                                                                           & \textbf{Type} \\ \hline
		Block Coverage           & \makecell[l]{each service needs to have exactly \\ one clinician that covers any given block}                                 & Hard          \\ \hline
		Weekend Coverage         & \makecell[l]{every weekend needs to have exactly \\ one clinician that covers it}                                              & Hard          \\ \hline
		Min/Max                  & \makecell[l]{for a given service, each clinician can only \\ work between the minimum and maximum \\ number of allowed blocks} & Hard          \\ \hline
		No Consecutive Blocks    & \makecell[l]{any clinician should not work \\ two consecutive blocks, across all services}                                    & Hard          \\ \hline
		No Consecutive Weekends  & \makecell[l]{any clinician should not work two consecutive \\ weekends}                                                        & Hard          \\ \hline
		Equal Weekends           & \makecell[l]{weekends should be equally distributed \\ between clinicians}                                                     & Hard          \\ \hline
		Equal Holidays           & \makecell[l]{long weekends should be equally distributed \\ between clinicians}                                                & Hard          \\ \hline
		Block Requests           & \makecell[l]{each clinician can request to be off service \\ during certain blocks throughout the year}                        & Soft          \\ \hline
		Weekend Requests         & \makecell[l]{each clinician can request to be off service \\ during certain weekends throughout the year}                      & Soft          \\ \hline
		Block-Weekend Adjacency  & \makecell[l]{the block and weekend assignments of a given \\ clinician should be adjacent}                                     & Soft
	\end{tabular}
	\caption{Summary of the constraints for the clinician scheduling problem}
	\label{tbl:constraint-summary}
\end{table}

### 2019-08-21 paper/paper.results.tex
Rename divisions -> services throughout
Also additional changes to results section following comments

\subsection{Infectious Diseases Division}  %main comment  = I think use past tense in writing
We use clinician time-off requests and minimum/maximum requirements from 2015-2018 as input data for the LP problem. Table \ref{tbl:2018-schedule-comparison} presents a the optimal schedule generated using the software as well as the manually-created schedule for data from 2018. The schedule is color-coded to distinguish between the different clinicians assigned.

\include{tbl/2018_schedule_comparison}
First, we evaluated the software-generated schedule by comparing it with the manually-created schedule. Specifically, we examined the adherence of each schedule to the constraints presented in table \ref{tbl:constraint-summary}. As shown in table \ref{tbl:constraints-comparison}, the optimal schedule generated using the software satisfied all mandatory constraints. In contrast, the manually-created schedule was not able to satisfy all mandatory constraints. In particular, we see that manual schedule assigned clinicians to multiple consecutive blocks for all four years. Moreover, the manual schedule did not have an equal distribution of weekends and holidays for all four years of data. Evaluating the objectives, we see that LP outperforms the manually created schedule in all four years, by accommodating almost all time-off requests and ensuring that weekends are always assigned close to blocks.

@@

\subsection{Simulations}
We then examined the properties of the LP problem using simulated data. Specifically, we examined the influence of expanding the following constraints on the runtime of the algorithm: larger number of  clinicians; clinicians with many time-off requests throughout the year. We then examined the effect of a longer time-horizon or a large number of services within a single department serviced by the same clinicians, on the runtime. \\

@@

Figure \ref{fig:avg-runtime-blocks} presents the average runtime over 10 runs for an increasing number of 2-week blocks $B = \{26, 52, 78, 104\}$ to simulate a department that generates schedules for a few years in advance. Although here was an increasing trend in the runtime, the solver was able to find an optimal schedule for a 4-year time horizon in a very reasonable amount of time. \\ % 'very reasonable amount of time" is vague. what is reasonable? better to give a range in value.

@@

The effect of increasing the number of divisions %do you mean services?
on the runtime of the program is shown in Table \ref{tbl:runtime-divisions-comparison} which presents the runtime for $D = \{1, 2, 3\}$ divisions and $C = \{10, 20, 30, 50\}$ clinicians in total across all divisions. For 2 or more divisions, a roster of more than 10 clinicians becomes impractical to schedule, as searching requires more than 24 hours. However, when relaxing the `No Consecutive Blocks' constraint, we see a great improvement in runtime for 2 and 3 divisions, and we are able to generate a schedule with upwards of 50 clinicians.  %what does division mean? do you mean services?

\include{tbl/runtime_divisions}

>>>>

\subsection{Infectious Diseases Division}  %main comment  = I think use past tense in writing
We used clinician time-off requests and minimum/maximum requirements from 2015-2018 as input data for the LP problem. Table \ref{tbl:2018-schedule-comparison} presents the optimal schedule generated using the software as well as the manually-created schedule for data from 2018. The schedule is color-coded to distinguish between the different clinicians assigned. \\

\include{tbl/2018_schedule_comparison}

First, we evaluated the software-generated schedule by comparing it with the manually-created schedule. Specifically, we examined the adherence of each schedule to the constraints presented in table \ref{tbl:constraint-summary}. As shown in table \ref{tbl:constraints-comparison}, the optimal schedule generated using the software satisfied all mandatory constraints. In contrast, the manually-created schedule was not able to satisfy all mandatory constraints. In particular, we see that the manual schedule assigned clinicians to multiple consecutive blocks in all four years. Moreover, the manual schedule did not have an equal distribution of weekends and holidays for all four years of data. Evaluating the objectives, we see that LP outperforms the manually created schedule in all four years, by accommodating almost all time-off requests and ensuring that weekends are always assigned close to blocks.

@@

\subsection{Simulations}
We then examined the properties of the LP problem using simulated data. Specifically, we examined the influence of expanding the following parameters on the runtime of the algorithm: number of clinicians; number of services offered in the division; amount of time-off requests per clinician throughout the year. We also examined the effect of a longer time-horizon on the runtime. \\

@@

Figure \ref{fig:avg-runtime-blocks} presents the average runtime over 10 runs for an increasing number of 2-week blocks $B = \{26, 52, 78, 104\}$ to simulate a department that generates schedules for a few years in advance. Although here was an increasing trend in the runtime, the solver was able to find an optimal schedule for all time horizons in under XX amount of time. \\ % 'very reasonable amount of time" is vague. what is reasonable? better to give a range in value.

@@

The effect of increasing the number of services %divisions %do you mean services?
on the runtime of the program is shown in Table \ref{tbl:runtime-services-clinicians-comparison}. We executed the algorithm for $S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in total across all services. For 2 or more services, a roster of more than XX clinicians becomes impractical to schedule, as searching requires more than 24 hours. However, when relaxing the `No Consecutive Blocks' constraint, we see a great improvement in runtime for 2 and 3 services, and we are able to generate a schedule with upwards of 50 clinicians in under XX amount of time.  %what does division mean? do you mean services?

\include{tbl/runtime_services_clinicians}
